Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification

### 2021-06-18 paper-revision/policar2020-tsne-embedding.tex
revision

@@

Batch effects are systematic biases between biological data sets caused by
technical factors in the data collection and preparation process. It has been
well documented that even small differences in the experimental setup of
cell-dissociation, handling protocols, library-preparation technologies, or
sequencing platforms can significantly affect the resulting gene-expression
measurements~\cite{Tung2017,Hicks2018}. When performing downstream comparative
analyses, batch effects may confound real biological variability and may
introduce spurious correlations, leading to misleading conclusions.

Due to their severity, numerous computational approaches have been proposed to
directlty remove batch effects when performing joint analysis on two or more data sets.
Batch effect removal is typically performed as a preprocessing step. Existing
approaches involve either modifying the original data matrix or finding a joint
lower-dimensional space, where batch-effects are removed. Current methods
broadly fall into two categories: 1) mutual nearest neighbor based approaches
aim to identify matching populations of cells across the data sets, using them
to either find and correct the data sets~\cite{Haghverdi2018} or directly
construct a batch-corrected k-nearest neighbor graph used in downstream
analyses~\cite{Park2018}. 2) embedding multiple data sets into a joint
lower-dimensional space, where batch effects are removed. Some of these
approaches opt for linear dimensionality-reduction methods such as
PCA~\cite{Korsunsky2019} or MultiCCA~\cite{Butler2018}, while others employ
non-linear techniques from deep learning~\cite{Li2020,Lopez2018} Still, other
approaches use a combination of the two~\cite{Stuart2019,Hie2019}.

Orthogonal approaches aimed at cell-type attempt to circumvent batch effects
altogether by identifying a subset of representative genes which act as markers
for a specific cell-type. Instead of considering the entire gene-expression
profile, which may be noisy and affected by batch effects, the idea is instead
to inspect the cells for the presence of a handful of genes, which could
adequately be used to determine the cell type. One such procedures is
scMap-Cluster, a consensus-based KNN method, tailored explicitly to scRNA-seq
gene-expression data~\cite{Kiselev2018}. scMap-Cluster uses three
correlation-based distance measures and uses a voting scheme to perform
classification. To identify novel cell types, scMap-Cluster heuristically
determines a distance threshold.

Broadly speaking, our approach falls into the second category, alongside
scMap-Cluster. The approach we describe uses standard single-cell pipelines
based on identifying representative genes. However, we emphasize that the
primary purpose of our approach is not classification, but the visualization of
the various cell-types. We can apply a KNN classifier to the resulting
visualizations to obtain accuracy estimates and compare our approach to other
classification methods. However, the classification aspect of our approach is
secondary: the primary purpose of t-SNE is to aid in scientists in exploratory
data analysis and help them better understand the underlying data landscape.

@@

\vspace{-3mm}
\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. The first data set in each pair
(Hrvatin \etal, Baron \etal, and Macosko \etal) was used as a reference.  We
relied on the quality control and annotations from the original publication
and report the number of cell types after preprocessinng.  The cell
annotations were made consistent to annotations from the Cell
Ontology~\cite{Bard2005}. Notice that different RNA sequencing protocols
were used to estimate gene expressions.}
\label{tab:data sets}
\end{table}

@@

Notice that we could simulate the split between reference and secondary data
sets using one data set only and perform cross-validation, however this type of
experiment would not incorporate batch effects. We want to remind the reader
that handling batch effects were central to our endeavour and that the
disregard of this effect could lead to overly-optimistic results and data
visualizations strikingly different from ours. For example, compare the
visualizations from Fig.~\ref{fig:batch_effect}.a and
Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and
\ref{fig:transform_pancreas}.b.

@@

\subsection{Constructing a reference embedding}

We use a number of additional, recently proposed modifications to enhance the t-SNE visualization of the reference data set.
Kobak and Linderman have shown that the global consistency of embeddings produced by
popular visualization algorithms are largely dependent on their initialization~\cite{Kobak2019a}. By utilizing PCA-based initialization, t-SNE is able to achieve
more meaningful layouts of the resulting clusters (Fig.~\ref{fig:multiscale}.b)
as opposed to using randomly initialized embeddings (Fig.~\ref{fig:multiscale}.a).
Another important extension is the use of multi-scale similarities, which, in addition
to considering short range interactions, also models wider point neighborhoods. Coupled
with PCA-based initializations, this produces even more meaningful visualizations where
clusters form interpretable structures. For instance, consider Fig.~\ref{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic neurons, corresponding to their
developmental origin, as discussed in Tasic \etal~\cite{Tasic2018}, while this division
is less apparent when using PCA-based initialization alone in
Fig.~\ref{fig:multiscale}.b.

We also observed the important role of gene selection in crafting the reference
embedding spaces. We found that when selecting an insufficient number of genes,
the resulting visualizations display overly-fragmented clusters. When the
selection is too broad and includes lowly expressed genes, the subclusters tend
to overlap. These effects can all be attributed to sparseness of the data sets
and may be intrinsic to single-cell data. In our studies, we found that
selection of 3,000 genes yields most informative visualizations
(Fig.~\ref{fig:gene_selection}).

@@

Fig.~\ref{fig:optimization} demonstrates a case where nearest neighbor-based 
positioning alone is insufficient. We construct a reference embedding using only
neurons from Hrvatin \etal~\cite{Hrvatin2018} (Fig.~\ref{fig:optimization}.a)
and use that to position neuronal
cells from the data set from Campbell \etal~\cite{Campbell2017}. We utilize the
weighted mean and median positions to initialize point positions from the secondary
data set, as shown in Figs.~\ref{fig:optimization}.b and \ref{fig:optimization}.c.
After initialization, we optimize point positions using the procedure described
above for 500 iterations. The resulting visualizations from both initializations
are visually very similar, indicating stable convergence. We show one of the resulting
visualizations in Fig.~\ref{fig:optimization}.b.

@@

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{optimization.pdf}
  \caption{Comparison of different initialization schemes for positioning new data
  points onto reference embeddings. {\bf (a)} We construct a reference embedding
  using only neuronal subtypes from Hrvatin \etal~\cite{Hrvatin2018}. {\bf (b)}
  We position neuronal cells from Campbell \etal~\cite{Campbell2017} using the
  median initizaliation scheme from Kobak and Berens~\cite{Kobak2019} and
  run optimization for 500 iterations. Compare the optimized embedding with the 
  initial median initialization {\bf (c)} or by using a simple weighted mean
  initialization {\bf (d)}.}
  \label{fig:optimization}
\end{figure}

\subsection{Our approach requires a complete reference set}

Our approach assumes that all cell types from the secondary data set are present
in the reference. Intuitively, using t-SNE in such a way is conceptually similar
to classification via $k$-nearest neighbor classifiers and is similarily
limited. The method may fail to reveal unseen cell types in the secondary data
set, likely positioning them arbitrarily close to unrelated clusters. In some
instances, unknown cell types may be sufficiently different from the reference
data that t-SNE will repel them from existing clusters. However, we caution that
this approach is unreliable and depends heavily on the chosen preprocessing
pipeline.

We illustrate this with Fig.~\ref{fig:incomplete_reference}, where we first fit
create a reference embedding containing only neuronal cells from Hrvatin
\etal~\cite{Hrvatin2018}. We then select only non-neuronal cells from Campbell
\etal~\cite{Campbell2017} and add them to the reference embedding in
Fig.~\ref{fig:incomplete_reference}.b. The non-neuronal cells from Hrvatin \etal
are scattered somewhat arbitrarily around several clusters in the reference
embedding. Interestingly, the secondary data points form a ``ring'' around one
of the clusters, indicating that these data points are very different from the
cells in this cluster. Notice also that the points from the secondary data set
exhibit little to no clustering and the different cell types seem to be mixed
amongst each other. We hypothesize that this effect is due primarily to the
single-cell preprocessing pipeline and not the limitations of our proceedure
itself, as the informative genes selected to create the reference neuronal
embedding likely do not differentiate supportive glial cells from the secondary
data set. This effect is similar to procedures such as scMap-Cluster, a consensus KNN
method, which heuristically determines a distance threshold to identify unknown
cell types~\cite{Kiselev2018}.

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{incomplete_reference-brain.pdf}
  \caption{A reference embedding must contain all the cell-types in the
  secondary embedding to produce reliable results. {\bf (a)} We construct a
  reference embedding containing only neuronal cells from Hrvatin
  \etal~\cite{Hrvatin2018}. {\bf (b)} We select only non-neuronal cells from
  Campbell \etal~\cite{Campbell2017} so that no overlap exists between the
  cell-types between the data sets. Conceptually, t-SNE behaves similarily to a
  KNN classifier, and places the non-neurnal cells to their the most similar
  points in the reference. In some instances, the non-neuronal cells are
  sufficiently different from the neuronal cells so that they are repelled from
  the reference clusters. Such behaviour results in the ``ring'' seen on the
  right hand side of the embedding.} \label{fig:incomplete_reference}
\end{figure}

@@

To quantitatively evaluate the predictive accuracy of the described procedure,
we fit $k$-nearest neighbors classifiers on each reference t-SNE embedding from
Figs.~\ref{fig:transform_brain}.a, \ref{fig:transform_pancreas}.a and
\ref{fig:transform_retina}.a and use them to predict the cell types for the
secondary data set embeddings from Figs.~\ref{fig:transform_brain}.b,
\ref{fig:transform_pancreas}.b and \ref{fig:transform_retina}.b. The accuracy
measures are reported in Table~\ref{tab:scores}. Our procedure of embedding new
data points into two-dimensional t-SNE plane results in similar accuracy to
approaches like random forests that use full compendium of cell-characterizing
features. The results indicate that positioning of new cells onto a cell
visualisation plane is not only indicative but also an accurate instrument for
cell type characterization.

@@

We compare our approach to two machine learning techniques, namely a $k$-nearest
neighbor classifier (KNN) and a random forest ensemble, and
scMap-Cluster~\cite{Kiselev2018}. For scMap-Cluster, we disable the distance
threshold heuristic for identifying novel cell types, as our secondary data sets
were chosen such that there is complete overlap between cell-types. For the two
machine learning approaches, we apply the typical single-cell preprocessing
pipeline described in Appendix~\ref{sec:sc-pipeline}, i.e., library-size
normalization, log-transformation, and select 1,000 most informative genes.
Similarily to scMap-Cluster, we use the cosine distance to find the 5 nearest
neighbors in the KNN model. We used 100 trees in the random forest ensemble. The
models were fit on the reference data set, and no hyper-parameter tuning was
performed.

@@

The aim of t-SNE is to construct embeddings, in which neighborhoods are
preserved, therefore it is unsurprising that the accuracy of our t-SNE based
approach is largely consistent with the $k$-nearest neighbors model. While our
approach is comparable to the other models in terms of accuracy, we emphasize
that the goal of t-SNE embeddings is to serve as visual aids in exploratory data
analysis. Therefore, it is surprising that our simple procedure performs
competitievly to specialized classification methods. Therefore, our procedure,
in addition to providing the end-user with a cell-type prediction, allows the
user to examine the low-dimensional embedding space, which may provide richer
insight and interpretation of the resulting predictions.

@@

Single-cell data sets suffer from high levels of technical noise and low capture
efficiency, resulting in sparse and noisy expression matrices~\cite{umi}. A
common occurence in these data sets is ``dropout'', where an expressed gene is
not measured, and its corresponding matrix entry is set to zero. To address this
problem, we use a specialized feature-selection method, which exploits the
mean-dropout relationship of expression counts as recently proposed by Kobak and
Berens~\cite{Kobak2019}. In this context, we will refer to all genes with matrix
entries set to zero as dropouts. Intuitively, if a gene has high mean expression
over all cells, but is detected in only a handful of them (i.e. has high mean
dropout rate), then this gene is likely specific to a specific cell-type and
will serve as a good feature for any subsequent analysis where we wish to
discriminate between cell types.

@@

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{tasic_multiscale.pdf}
  \caption{A comparison of standard and multi-scale t-SNE on data from the
  mouse neocortex~\cite{Tasic2018}. {\bf (a)} Standard t-SNE using random 
  initialization places clusters arbitrarily. The resulting clustering structure is not globally consistent,
  as clusters of the same type of cells are dispersed throughout the landscape. Non-Neuronal clusters, for instance, are mixed with clusters of GABAergic and Glutamaergic neurons.
  {\bf (b)} By utilizing a globally consistent initialization for t-SNE, the clusters
  are organized in a more meaningful layout, where clusters of cells of the same type appear closer together.
  {\bf (c)} Augmenting t-SNE with multi-scale
  similarities and using proper initialization provides a more meaningful
  layout of the clusters. Non-Neuronal and Endothelial cell types are now placed in
  the same region of the embedding. There are two clear sub-groups of GABAergic
  neurons corresponding to their developmental origins, which was not as apparent
  when using clever initialization alone.}
  \label{fig:multiscale}
\end{figure}

>>>>

Our contributions are two-fold: 
\begin{enumerate}
  \item We introduce a theoretically-grounded extension of the t-SNE
   visualization algorithm that supports embedding new data points into
   existing reference visualizations. Our extension is readily incorporated
   into existing approximation schemes, enabling its applications to large
   data sets. We show that optimization using the default t-SNE parameters is
   highly unstable and proposes parameter values leading to stable
   convergence.

  \item We show that the proposed t-SNE extensions can mitigate batch effects
   in the data sets and demonstrate this feature in treating single-cell gene
   expression data.
\end{enumerate}

@@

Batch effects are systematic biases between biological data sets caused by
technical factors in the data collection and preparation process. It has been
well documented that even small differences in the experimental setup of
cell-dissociation, handling protocols, library-preparation technologies, or
sequencing platforms can significantly affect the resulting gene-expression
measurements~\cite{Tung2017,Hicks2018}. When performing downstream
comparative analyses, batch effects may confound real biological variability
and introduce spurious correlations, leading to misleading conclusions.

Due to their severity, numerous computational approaches have been proposed to
directly remove batch effects when performing joint analysis on two or more
data sets. Batch effect removal is typically performed as a preprocessing
step. Existing approaches involve either modifying the original data matrix
or finding a joint lower-dimensional space, where batch effects are removed.
Current methods broadly fall into two categories:
\begin{enumerate}
\item mutual nearest neighbor-based approaches aim to identify matching
 populations of cells across the data sets, using them to either find and
 correct the data sets~\cite{Haghverdi2018} or directly construct a
 batch-corrected k-nearest neighbor graph used in downstream analyses~\cite
 {Park2018}.
\item embedding multiple data sets into a joint lower-dimensional space, where
 batch effects are removed. Some of these approaches opt for linear
 dimensionality-reduction methods such as PCA~\cite{Korsunsky2019} or
 MultiCCA~\cite{Butler2018}, while others employ non-linear techniques from
 deep learning~\cite{Li2020,Lopez2018}. Still, other approaches use a
 combination of the two~\cite{Stuart2019,Hie2019}.
\end{enumerate}

Besides computational techniques, approaches for the removal of batch effects
can also use domain knowledge. For example, in the analysis of single-cell
gene expression data, these approaches act on a subset of representative
marker genes for a specific cell type. Instead of considering the entire
gene-expression profile, which may be noisy and affected by batch effects,
the idea is to profile the cells with a handful of genes that can
collectively determine the cell type. One such procedure is scMap-Cluster, a
consensus-based $k$-nearest neighbor method tailored explicitly to scRNA-seq
gene-expression data~\cite{Kiselev2018}. scMap-Cluster uses three
correlation-based distance measures and uses a voting scheme to perform
classification. To identify novel cell types, scMap-Cluster heuristically
determines a distance threshold.

Our approach to batch effect removal falls into the second category, as we
lose the batch effects through dimensionality reduction. Alongside
scMap-Cluster, we also benefit from a standard single-cell data preprocessing
pipeline that profiles the cells with representative genes. Unlike other
batch effect removal procedures, the primary purpose of our approach is not
classification but the visualization of the various cell-types. If required,
we can apply a $k$-nearest neighbor classifier to the resulting
visualizations to obtain accuracy estimates and compare our approach to other
classification methods. However, the classification aspect of our approach is
secondary: the primary purpose of t-SNE is to aid in scientists in
exploratory data analysis and help them better understand the underlying data
landscape.

@@

\vspace{-3mm}
\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. The first data set in each pair
(Hrvatin \etal, Baron \etal, and Macosko \etal) was used as a reference.  We
relied on the quality control and annotations from the original publication
and report the number of cell types after prepossessing.  The cell
annotations were made consistent to annotations from the Cell
Ontology~\cite{Bard2005}. Notice that different RNA sequencing protocols
were used to estimate gene expressions.}
\label{tab:data sets}
\end{table}

@@

Notice that we could simulate the split between reference and secondary data
sets using one data set only and perform cross-validation, however this type of
experiment would not incorporate batch effects. We want to remind the reader
that handling batch effects were central to our endeavor and that the
disregard of this effect could lead to overly-optimistic results and data
visualizations strikingly different from ours. For example, compare the
visualizations from Fig.~\ref{fig:batch_effect}.a and
Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and
\ref{fig:transform_pancreas}.b.

@@

\subsection{Construction of a reference embedding}

We use a number of additional, recently proposed modifications to enhance the
t-SNE visualization of the reference data set. Kobak and Linderman have shown
that the global consistency of embeddings produced by popular visualization
algorithms are largely dependent on their initialization~\cite
{Kobak2019a}. By utilizing PCA-based initialization, t-SNE is able to achieve
more meaningful layouts of the resulting clusters (Fig.~\ref
{fig:multiscale}.b) as opposed to using randomly initialized embeddings
(Fig.~\ref{fig:multiscale}.a). Another important extension is the use of
multi-scale similarities, which, in addition to considering short range
interactions, also models wider point neighborhoods. Coupled with PCA-based
initialization, this produces even more meaningful visualizations where
clusters form interpretable structures. For instance, consider Fig.~\ref
{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic
neurons, corresponding to their developmental origin, as discussed in
Tasic \etal~\cite{Tasic2018}, while this division is less apparent when using
PCA-based initialization alone in Fig.~\ref{fig:multiscale}.b.

We also observed the important role of gene selection in crafting the
reference embedding spaces. We found that when selecting an insufficient
number of genes, the resulting visualizations display overly-fragmented
clusters. When the selection is too broad and includes lowly expressed genes,
the subclusters tend to overlap. These effects can all be attributed to
sparseness of the data sets and may be intrinsic to single-cell data. In our
studies, we found that selection of 3,000 genes yields most informative
visualizations(Fig.~\ref{fig:gene_selection}).

@@

Fig.~\ref{fig:optimization} demonstrates a case where nearest neighbor-based
positioning alone is insufficient. We construct a reference embedding using
only neurons from Hrvatin \etal~\cite{Hrvatin2018} (Fig.~\ref
{fig:optimization}.a) and use that to position neuronal cells from the data
set from Campbell \etal~\cite{Campbell2017}. We utilize the weighted mean and
median positions to initialize point positions from the secondary data set,
as shown in Figs.~\ref{fig:optimization}.b and \ref
{fig:optimization}.c. After initialization, we optimize point positions using
the procedure described above for 500 iterations. The resulting
visualizations from both initializations are visually very similar,
indicating stable convergence. We show one of the resulting visualizations in
Fig.~\ref{fig:optimization}.b.

@@

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{optimization.pdf}
  \caption{Comparison of different initialization schemes for positioning new
   data points onto reference embeddings. {\bf (a)} We construct a reference
   embedding using only neuronal subtypes from Hrvatin \etal~\cite
   {Hrvatin2018}. {\bf (b)} We position neuronal cells from
   Campbell \etal~\cite{Campbell2017} using the median initialization scheme
   from Kobak and Berens~\cite{Kobak2019} and run optimization for 500
   iterations. Compare the optimized embedding with the initial median
   initialization {\bf (c)} or by using a simple weighted mean
   initialization {\bf (d)}.}
  \label{fig:optimization}
\end{figure}

\subsection{On requirement of a complete reference set}

Our approach assumes that all cell types from the secondary data set are
present in the reference. Intuitively, using t-SNE in such a way is
conceptually similar to classification via $k$-nearest neighbor classifiers
and is similarly limited. The method may fail to reveal unseen cell types in
the secondary data set, likely positioning them arbitrarily close to
unrelated clusters. In some instances, unknown cell types may be sufficiently
different from the reference data that t-SNE will repel them from existing
clusters. However, we caution that this approach is unreliable and depends
heavily on the chosen preprocessing pipeline.

We illustrate this with Fig.~\ref{fig:incomplete_reference}, where we first fit
create a reference embedding containing only neuronal cells from Hrvatin
\etal~\cite{Hrvatin2018}. We then select only non-neuronal cells from Campbell
\etal~\cite{Campbell2017} and add them to the reference embedding in Fig.~\ref
{fig:incomplete_reference}.b. The non-neuronal cells from Hrvatin \etal are
scattered somewhat arbitrarily around several clusters in the reference
embedding. Interestingly, the secondary data points form a ``ring'' around
one of the clusters, indicating that these data points are very different
from the cells in this cluster. Notice also that the points from the
secondary data set exhibit little to no clustering and the different cell
types seem to be mixed among each other. We hypothesize that this effect is
due primarily to the single-cell preprocessing pipeline and not the
limitations of our procedure itself, as the informative genes selected to
create the reference neuronal embedding likely do not differentiate
supportive glial cells from the secondary data set. This effect is similar
to procedures such as scMap-Cluster, a consensus $k$-nearest neighbor 
method, which heuristically determines a distance threshold to identify
unknown cell types~\cite{Kiselev2018}.

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{incomplete_reference-brain.pdf}
  \caption{A reference embedding must contain all the cell types in the secondary embedding to produce reliable results. {\bf (a)} We construct a reference embedding containing only neuronal cells from Hrvatin \etal~\cite {Hrvatin2018}. {\bf (b)} We select only non-neuronal cells from Campbell \etal~\cite{Campbell2017} so that no overlap exists between the cell types between the data sets. Conceptually, t-SNE behaves similarly to a $k$-NN classifier and places the non-neuronal cells to their most similar points in the reference. In some instances, the non-neuronal cells are sufficiently different from the neuronal cells so that they are repelled from the reference clusters. Such behavior results in the ``ring'' seen on the right-hand side of the embedding.} \label{fig:incomplete_reference}
\end{figure}

@@

To quantitatively evaluate the predictive accuracy of the described procedure,
we fit $k$-nearest neighbors classifiers on each reference t-SNE embedding from
Figs.~\ref{fig:transform_brain}.a, \ref{fig:transform_pancreas}.a and
\ref{fig:transform_retina}.a and use them to predict the cell types for the
secondary data set embeddings from Figs.~\ref{fig:transform_brain}.b,
\ref{fig:transform_pancreas}.b and \ref{fig:transform_retina}.b. The accuracy
measures are reported in Table~\ref{tab:scores}. Our procedure of embedding new
data points into two-dimensional t-SNE plane results in similar accuracy to
approaches like random forests that use full compendium of cell-characterizing
features. The results indicate that positioning of new cells onto a cell
visualization plane is not only indicative but also an accurate instrument for
cell type characterization.

@@

We compare our approach to two machine learning techniques, namely a $k$-nearest
neighbor classifier (KNN) and a random forest ensemble, and
scMap-Cluster~\cite{Kiselev2018}. For scMap-Cluster, we disable the distance
threshold heuristic for identifying novel cell types, as our secondary data sets
were chosen such that there is complete overlap between cell-types. For the two
machine learning approaches, we apply the typical single-cell preprocessing
pipeline described in Appendix~\ref{sec:sc-pipeline}, i.e., library-size
normalization, log-transformation, and select 1,000 most informative genes.
Similarly to scMap-Cluster, we use the cosine distance to find the 5 nearest
neighbors in the KNN model. We used 100 trees in the random forest ensemble. The
models were fit on the reference data set, and no hyper-parameter tuning was
performed.

@@

The aim of t-SNE is to construct embeddings, in which neighborhoods are
preserved, therefore it is unsurprising that the accuracy of our t-SNE based
approach is largely consistent with the $k$-nearest neighbors model. While our
approach is comparable to the other models in terms of accuracy, we emphasize
that the goal of t-SNE embeddings is to serve as visual aids in exploratory data
analysis. Therefore, it is surprising that our simple procedure performs
competitively to specialized classification methods. Therefore, our procedure,
in addition to providing the end-user with a cell-type prediction, allows the
user to examine the low-dimensional embedding space, which may provide richer
insight and interpretation of the resulting predictions.

@@

Single-cell data sets suffer from high levels of technical noise and low capture
efficiency, resulting in sparse and noisy expression matrices~\cite{umi}. A
common occurrence in these data sets is ``dropout'', where an expressed gene is
not measured, and its corresponding matrix entry is set to zero. To address this
problem, we use a specialized feature-selection method, which exploits the
mean-dropout relationship of expression counts as recently proposed by Kobak and
Berens~\cite{Kobak2019}. In this context, we will refer to all genes with matrix
entries set to zero as dropouts. Intuitively, if a gene has high mean expression
over all cells, but is detected in only a handful of them (i.e. has high mean
dropout rate), then this gene is likely specific to a specific cell-type and
will serve as a good feature for any subsequent analysis where we wish to
discriminate between cell types.

@@

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{tasic_multiscale.pdf}
  \caption{A comparison of standard and multi-scale t-SNE on data from the
   mouse neocortex~\cite{Tasic2018}. {\bf (a)} Standard t-SNE using random
   initialization places clusters arbitrarily. The resulting clustering
   structure is not globally consistent, as clusters of the same type of
   cells are dispersed throughout the landscape. Non-Neuronal clusters, for
   instance, are mixed with clusters of GABAergic and Glutamaergic neurons.
   {\bf (b)} By utilizing a globally consistent initialization for t-SNE, the
   clusters are organized in a more meaningful layout, where clusters of
   cells of the same type appear closer together.{\bf (c)} Augmenting t-SNE
   with multi-scale similarities and using proper initialization provides a
   more meaningful layout of the clusters. Non-Neuronal and Endothelial cell
   types are now placed in the same region of the embedding. There are two
   clear sub-groups of GABAergic neurons corresponding to their developmental
   origins, which was not as apparent when using clever initialization
   alone.}
  \label{fig:multiscale}
\end{figure}
